---
title: "Model Selection"
subtitle: "GPA Example" 
output:
  html_document:
    toc: yes
---
```{r setup2, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
packages_needed <- c("ggplot2",
                     "patchwork",
                     "MuMIn"
                     )
pk_to_install <- packages_needed [!( packages_needed %in% rownames(installed.packages())  )]
if(length(pk_to_install)>0 ){
  install.packages(pk_to_install,repos="http://cran.r-project.org")
}
#lapply(packages_needed, require, character.only = TRUE)
library(ggplot2)
library(patchwork)
library(MuMIn)
```

```{r, include=FALSE}
# change na. action
options(na.action = "na.fail")
```
GPA example from Graybill and Iyer (1994); worked in Burnham and Anderson (2002) p. 226
SAT scores for Math and Verbal sections; High School GPAs for Math and English
```{r}
GPA_data <- read.csv("data/GPA.txt", sep="")
```

```{r fig.height=8, fig.width=8}
#check for co-linearity
pairs(GPA_data[,3:6], lower.panel = NULL)
```
```{r look at the correlation matrix}
correlation.matrix <- cor(GPA_data[,3:6])
round(correlation.matrix, 2)
```

No obvious colinearity among variables. Anything above ~0.5 we should look at, but OK for now.

```{r}
# fit the full model with all parameters (sometimes called 'global')
GPA.all.parms<-lm(GPA ~ SATmath + SATverbal + HSmath + HSverbal, data = GPA_data)

# the dredge function fits all combinations
# of the variables in the GPA.all.parms model fit above
results<-MuMIn::dredge(GPA.all.parms) # note use of 'MuMIn' package
results
```

```{r}
# grab best supported models
subset(results, delta <5)
```
Models 14 and 13 have delta AICc <2. Model 14 is about 1.69 times more likely than Model 13 (0.485/0.287=1.69).
```{r}
# calculate variable importance weights
MuMIn::importance(results)
```
Tells us that HSverbal is not really a useful predictor, and it is absent in all of the 3 top models.

Only fit models without HSverbal.

```{r}
results<-dredge(GPA.all.parms, subset= !(HSverbal))
results
```
Again, the model selection results without including HSverbal are the same as those including HSverbal; Models 13 and 14 have about equal explanatory power (delta AICc <2). The inclusion of HS math does not explain additional variation above what is explained by the other three variables; inclusion of HSmath means a 1 parameter "penalty". Model 14 is about 1.69 times more likely than Model 13 (0.532/0.315=1.69). This re-scaled result is the same as before. 

```{r make a figure using best-fit model}
#trying to represent combinations of our three most important variables on predicting GPA
p3 <- ggplot(GPA_data, aes(SATmath, GPA, colour = SATverbal)) + 
  geom_point(size=3) +
  geom_smooth(method="lm")

p4 <- ggplot(GPA_data, aes(SATverbal, GPA, colour = HSmath)) + 
  geom_point(size=3) +
  geom_smooth(method="lm") +
  scale_colour_gradientn(colours = heat.colors(10, rev=TRUE))
                     
p3+p4 #uses 'patchwork' library to put plots next to each other
```

"Dredging" used here is just an example for comparing models. Dregding is known to prefer over-complex models.
https://dynamicecology.wordpress.com/2015/05/21/why-aic-appeals-to-ecologists-lowest-instincts/
